**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**

### 2019-07-22 submission/manuscript.Rmd
update manuscript for Pat

Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce due to the flawed methods that call the validity of developed ML models into question. Furthermore, the use of black box ML models has hindered the validation of microbial biomarkers. To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained much faster (12 min). In this study, we established standards for the development of modeling pipelines for microbiome-associated ML models. Additionally, we showed that ML models should be chosen based on expectations of predictive performance, interpretability and available computational resources.

@@

Moreover, there is a lack of discussion on why a particular ML model is utilized and a lack of emphasis on the strengths and weaknesses of it. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting (XGBoost) and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018]. These models are also called black box ML models because they are not inherently interpretable and require posthoc explanations to determine the feature importances in making a prediction. These explanations can be misleading and at times unreliable when making high-stake decisions about patient health [@rudin_please_2018]. The models we develop for healthcare, both to predict disease and to understand underlying reasons behind that prediction, should be transparent and accountable [@rudin_optimized_2018].

@@

To highlight the differences between the two interpretation methods, we used permutation importance to interpret linear models as well [Figure S3]. L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in feature rankings based on weights [Figure 3] and permutation importance [Figure S3]. Similarly, L2-regularized SVM and L2-regularized logistic regression picked out some of the same OTUs in both interpretation methods, OTU00659 and OTU00012, respectively. However, for all the linear models, the rankings of these features were different due to the collinearity in microbial communities.

@@

We compared the training times of the seven ML models. As the complexity of a ML model [Table S1] and the number of tuned hyperparameter settings increased [Figures S1-S2], its training times increased as well [Figure 5]. Linear models trained faster than non-linear models. L1 and L2 SVM with linear kernel and L2-regularized logistic regression had the shortest training times with `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively. Whereas, a decision tree, SVM with radial basis function kernel, random forest and XGBoost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 5].

@@

In this study we established a rigorous ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome. We built on others' work in the microbiome field to set-up standards for developing and evaluating ML models for microbiome data [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019]. First, we used a held-out test set to illustrate the difference between cross-validation and testing AUROC values. When the difference between cross-validation and test performance is low, this suggest the models are not overfit and that they will perform similar with similar data. In all seven models, the difference between median cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models are generalizable and can be used to test similar new data. Second, we performed the initial 80%-20% random datasplit 100 times in our ML pipeline. The randomization of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop robust ML models and to report reliable performance metrics. Depending on how the data is split, there is the chance of being overoptimistic about the predictive performance of a model. In our study, we showed that there was variability in AUROC values between different random data-splits in each of the models we tested. Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average between different data-splits. Third, we used the AUROC metric in our study to evaluate the predictive performance of the ML models. AUROC is always random at the value 0.5 and is a robust metric when a dataset is imbalanced. We also performed a full grid search for hyperparameter settings when building a ML model. Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new dataset used to generate a model. In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.

In this study, we benchmarked seven ML models with different classification algorithms to show that we should use ML models based on the goal of the study and our expectations of predictive performance, interpretability and computational burden. Microbiome studies use ML models with a classification task to learn the training data to assign labels, such as healthy or not to new data, but also to learn which features are important to discriminate between labels [@zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018]. Our results show that if the goal of a study is to learn the ecology behind a disease and to identify microbial biomarkers, we could create ML models that are inherently interpretable without losing predictive power. In terms of predictive performance, random forest model had the best testing AUROC values compared to the other 6 models. However, the second best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest. In terms of interpretability, random forest was a complex ML model and could only be explained using methods such as permutation importance. On the other hand, L2-regularized logistic regression was easier to interpret (i.e. ranking absolute regression coefficients of the trained model).

Even with interpretable models such as L2-regularized logistic regression, we need to be careful with how we treat the information we get from the models. In this study we used two different methods to interpret our linear models; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance. We observed differences in the OTU rankings between the two interpretation methods due to collinearity in the dataset. Collinearity in a microbial dataset is when one OTU is highly correlated with another OTU which causes feature weights to not be unique. The feature weights of correlated OTUs are influenced by one another which makes it difficult to interpret the ML model. To avoid misinterpreting the models we should use the highly ranked correlated OTUs to generate hypotheses about the ecology of the disease and test them with follow-up experiments.

There are other criteria when choosing ML models such as the computational burden of developing it and the sample size. In terms of computational burden, random forest model trained each data-split in 83.2 hours whereas L2-regularized logistic regression trained in 12 minutes. The generalization performance of ML models depends on sample size. The more complex the model, the more data it will need. The dataset we used for our study had 490 samples, however microbiome studies that have smaller sample sizes would benefit from using less complex models such as L2-regularized logistic regression.

This study highlights the need to make educated choices at every step of developing a ML model with microbiome data. Model selection should be done with a solid understanding of model complexity and interpretability, rigorous ML pipelines should be built with cross-validation for hyperparameter tuning and with a held-out test set for evaluating predictive performance and models should be interpreted while considering collinearity in datasets. The right methods will help us achieve the level of validity and accountability we want from models built for patient health.

@@

\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\seperating hyperplane that discriminates between labels.} & Linear & Interpratable & $^{37}$\\ \hline

\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal gaussian \\seperating hyperplane that discriminates between labels.} & Non-linear & Explainable$^*$ & $^{38}$\\ \hline
Decision tree & \makecell[l]{A classifier that sorts samples down from the 
root to the \\leaf node where an attribute is tested to discriminate \\between labels} & Non-linear & Interpretable & $^{39}$\\ \hline

@@

\makecell[l]{Model \\ selection} & \makecell[l]{Do we know the \\ candidate algorithms \\ for the ML problem? \\} & \makecell[l]{Do we know our \\ computational resources \\ to fully train each model?} & \makecell[l]{How much interpretibility \\ does the problem need? \\ How much each \\ candidate algorithm \\ can provide?} \\ \hline

>>>>

Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce due to the flawed methods that call the validity of developed models into question. Furthermore, the use of black box ML models has hindered the validation of microbial biomarkers. To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained much faster (12 min). In this study, we established standards for the development of modeling pipelines for microbiome-associated ML models. Additionally, we showed that ML models should be chosen based on expectations of predictive performance, interpretability and available computational resources.

@@

Moreover, there is a lack of discussion on why a particular ML model is utilized and a lack of emphasis on the strengths and weaknesses of it. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting (XGBoost) and neural networks without a discussion on if and how much model interpretability is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018]. These models are also called black box ML models because they are not inherently interpretable and require posthoc explanations to determine the feature importances in making a prediction. These explanations can be misleading and at times unreliable when making high-stake decisions about patient health [@rudin_please_2018]. The models we develop for healthcare, both to predict disease and to understand underlying reasons behind that prediction, should be transparent and accountable [@rudin_optimized_2018].

@@

To highlight the differences between the two interpretation methods, we used permutation importance to interpret linear models as well [Figure S3]. L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in feature rankings based on weights [Figure 3] and permutation importance [Figure S3]. Similarly, L2-regularized SVM and L2-regularized logistic regression picked out some of the same OTUs in both interpretation methods, OTU00659 and OTU00012, respectively. However, for all the linear models, the rankings of these features were different due to the collinearity in microbial communities. Collinearity in a microbial dataset is when one OTU is dependant on another OTU. The feature weights of correlated OTUs are influenced by one another which makes it difficult to interpret models using feature weights. Our interpretation of L2-regularized logistic regression based on feature weight rankings, showed that an OTU that belongs to *Lachnospiraceae* (OTU00056) was not among the most important OTUs in making a prediction [Figure 3C]. However, it was picked out as important when the same model was interpreted using permutation importance [Figure S3C]. This is due to collinearity in the dataset, where *Lachnospiraceae* (OTU00056) is significantly correlated with *Feacalibacterium* (OTU00015) which has the highest ranked feature weights [Figure 3C]. We need to investigate the relationships among features to identify the true underlying factors when making a prediction.

@@

We compared the training times of the seven ML models. As the complexity of a ML model and the number of tuned hyperparameter settings increased [Figures S1-S2], its training times increased as well [Figure 5]. Linear models trained faster than non-linear models. L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression had the shortest training times with `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively. Whereas, a decision tree, SVM with radial basis function kernel, random forest and XGBoost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 5].

@@

Microbiome studies use ML models, often with a classification task, to predict a disease but also to learn which microbes are indicators of that disease [@zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018]. Achieving either of these tasks have far-reaching impact on human health, however ML as a tool in microbiome studies is still at its infancy. A framework is needed to develop rigorous ML models, to identify and overcome potential pitfalls. Previous studies generated workflows to allow ML to be widely used by the microbiome researchers [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019]. This study sets-up standards for developing and evaluating rigorous ML models for microbiome data [Table 1].

We benchmarked seven ML models with different classification algorithms to show that a clearly defined ML problem that is based on the goal of the microbiome study should inform our model selection. Our results showed that if the goal of a study is to learn the ecology behind a disease and to identify microbial biomarkers, we can create ML models that are inherently interpretable and easily trained without losing predictive power. In terms of predictive performance, random forest model had the best testing AUROC values compared to the other six models. However, the second best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest. While random forest took 83.2 hours to train, L2-regularized logistic regression trained in 12 minutes. In terms of interpretability, random forest was a more complex ML model and could only be explained using methods such as permutation importance. On the other hand, L2-regularized logistic regression was easier to interpret by ranking absolute feature weights of the trained model.

Even with interpretable models such as L2-regularized logistic regression, there are potential pitfalls when it comes to identifying biomarkers of a disease. As domain experts, we know that human-associated microbial communities have complex correlation structures that creates collinearity in the dataset. Collinearity is a severe problem and needs to be addressed to reliably interpret ML models [@dormann_collinearity:_2013]. In this study we used two different methods to interpret our linear models; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance. We observed differences in the OTU rankings between the two interpretation methods due to collinearity in the dataset. To avoid misinterpreting the models, once we identify the most important microbes, we should check for their relationships with other microbes as well. These relationships will help us generate new hypotheses about the ecology of the disease. These hypotheses needs to be tested with follow-up experiments to identify the true biomarkers of a disease.

In this study, we also established a rigorous ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome. First of all, we used a held-out test set to illustrate the difference between cross-validation and testing AUROC values. When the difference between cross-validation and test performance is low, this suggest the models are not overfit and that they will perform similar with similar data. In all the models, the difference between median cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models are generalizable and can be used to test similar new data. Furthermore, we performed the initial 80%-20% random datasplit 100 times in our ML pipeline. Depending on how the data is split, there is the chance of being overoptimistic about the predictive performance of a model. We showed that there was variability in AUROC values between different random data-splits in each of the models we tested. Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average between different data-splits. The randomization and resampling of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop robust ML models and to report reliable performance metrics. We also performed a full grid search for hyperparameter settings when training our ML models. Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new dataset used to generate a model. In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default. And finally, we used the AUROC metric in our study to evaluate the predictive performance of the ML models. AUROC is always random at the value 0.5 and is a robust metric when a dataset is imbalanced.

We used a balanced CRC dataset to develop ML models with a binary classification task. We did not evaluate multicategory classification methods or regression analyses to predict non-binary outcomes. However, the principles highlighted throughout this study [Table 1] apply to all ML modeling tasks with microbiome data. The models we built were generalizable despite the high number of features microbiome datasets usually have. The generalization performance of ML models depends on sample size. The more complex the model, the more data it will need. Our dataset had 490 samples, however microbiome studies that have smaller sample sizes would benefit from using less complex models. Our analysis was limited to shallow learning methods and did not explore deep learning methods such as neural networks. Microbiome datasets often suffer from having high dimensionality but low sample sizes which makes deep learning models prone to overfitting. There are studies that address overcoming these challenges in biomedical datasets [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018], however sudies that estalish frameworks with microbiome data are lacking. This would be an interesting direction for future work in microbiome studies.

This study highlighted the need to make educated choices at every step of developing a ML model with microbiome data. Model selection should be done with a solid understanding of model complexity and interpretability, rigorous ML pipelines should be built with cross-validation for hyperparameter tuning and with a held-out test set for evaluating predictive performance and models should be interpreted while considering collinearity in datasets. The right methods will help us achieve the level of validity and accountability we want from models built for patient health.

@@

\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear & Interpretable & $^{37}$\\ \hline

\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal Gaussian \\separating hyperplane that discriminates between labels.} & Non-linear & Explainable$^*$ & $^{38}$\\ \hline
Decision tree & \makecell[l]{A classifier that sorts samples down from the 
root to the \\leaf node where an attribute is tested to discriminate \\between labels} & Non-linear & Interpretable & $^{39}$\\ \hline

@@

\makecell[l]{Model \\ selection} & \makecell[l]{Do we know the \\ candidate algorithms \\ for the ML problem? \\} & \makecell[l]{Do we know our \\ computational resources \\ to fully train each model?} & \makecell[l]{How much interpretability \\ does the problem need? \\ How much each \\ candidate algorithm \\ can provide?} \\ \hline
